---
title: ""
subtitle: "R from zero"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
---

#### prep

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8, warning=FALSE, message=FALSE, cache=TRUE, error=T)
```

```{r echo=F, include=F}
library(tidyverse)
library(magrittr)
library(psych)
library(foreign)
library(sjlabelled)
library(sjmisc)
library(HH)
library(sjPlot)
library(readxl)
library(DescTools)
library(janitor)
library(skimr)
library(xlsx)
```

```{r include=F}

options(digits=3, scipen=6)

# vertical and horizontal lines
theme_set(theme_bw() + theme(panel.grid.minor.x=element_blank(),
                             panel.grid.minor.y=element_blank(),
                             plot.title=element_text(face="bold",size=20, hjust=.5),
                             axis.title=element_text(size=18),
                             axis.text=element_text(size=16),
                             legend.text=element_text(size=16),
                             strip.text=element_text(size=16)))

setwd("C:/Egnyte/Private/dkillian/R from zero")

```

## About R

R represents an implementation of what is known as object-oriented programming. 

This means that information can be stored as 'objects', and R code can act on these objects. 

We create an object by using the assignment operator: <- 

```{r}
a <- 4
a
```

```{r}
b <- 1:3
b
```

```{r}
c <- letters[1:3]
c
```

```{r}

```

```{r}

```

## Data types

There are four primary types of data in r: 

* character
* integer
* numeric
* logical

### Characters

Character data holds string values. A string is information contained within a set of quotation marks. 

```{r}
char <- c("This is a string", "This is also a string", "Is this a string?")
char

```

```{r}
c
```

```{r}
d <- c(3,2,a,b)
d
```
```{r}
e <- c(3,2,z,x)
```
R throws an error because z is not a string (there are no quotations around it) and z is not an object in the environment. Therefore, R cannot find z and cannot complete the operation. 

```{r}
e <- c(3,2,"z","x")
e
```

R coerced the integers into characters so that all elements would be of the same type. 

### integers

Integers are real-valued numbers. 

We create an integer using the suffix L. Otherwise, it will be numeric even if displayed as an integer. 

```{r}
str(b)
```

```{r}

f <- c(1,2,3)
str(f)

```

```{r}
g <- c(1L, 2L, 3L)
str(g)

```

```{r}
toni <- c(1,5,3,2,5)
toni
```

```{r}
toni2 <- c(1,4,1:4,3,1)
toni2
```


### numbers

Numbers are an infinite collection of values along the real number line.  

```{r}
h <- c(1.1, 2.2, 3.3)
str(h)


```

### logicals

Logical data simply indicates whether or not a piece of information satisfies some condition. Logical data only takes on two values: True (T) or False (F). 

```{r}
a>2
```

```{r}
b>2
```

```{r}
c > 2
```

For some reason, R believes that strings are greater than 2. Be careful what you ask!

```{r}
sum(b<3)
```

#### Real world application: checking indices

```{r}

index <- 1:100

var1 <- rnorm(100, 25,5)
var2 <- rnorm(100, 50,10)


dat1 <- data.frame(index, var1) 
str(dat1)

```

```{r}
dat2 <- data.frame(index, var2) 
str(dat2)
```

```{r}
dat1$index == dat2$index
```

```{r}
sum(dat1$index==dat2$index)
```

```{r}
dat2 <- arrange(dat2, desc(dat2$index))
head(dat2)
```

```{r}
sum(dat1$index==dat2$index)
```

## Data structures

R stores data in four primary structures: vector, list, matrix, and data frame. 

### vectors

A vector is a list of elements, where all elements must be of the same type. 

```{r}
a
```

```{r}
b
```

```{r}
a + b
```

```{r}
d
```

```{r}
b + d
```

Try to assign data of different type to a vector: 

```{r}
i <- c(2, "a")
str(i)
```

The result is coercion into a single data type. In this case, the 2 was converted to a character. 

### lists

A list is a vector that can hold multiple data types. 

```{r}
j <- as.list(c(2, "a"))
str(j)
```

```{r}
j[[1]]
```

```{r}
j[2]
```

```{r}
k <- list(1:10)
k
```

```{r}
l <- list(letters[1:10])
l
```

```{r}
m <- list(k,l)
m[1]
```


#### matrices

A matrix is a two dimensional collection of vectors (rows and columns). 

```{r}
n <- matrix(1:25, ncol=5)
n
```

```{r}
o <- matrix(1:25, ncol=5,byrow=T)
o
```

As with vectors, the data in a matrix must all be of the same type. 

Matrices can be used in calculations. 

```{r}
n + o
```

```{r}
n * o
```

```{r}
n %*% o
```

```{r}
eigen(g)
```


### data frames

A data frame is a matrix that can hold different types of data across columns (but not within columns). 

Another way to think of a data frame is that it is a collection of lists, where each column is an element of the list. 

A data frame is the object type we most often use for analysis. 

```{r}
dat_ex <- data.frame(a=rnorm(100, 25,10), b=1:100, c=rep(letters[1:25],4))
dat_ex
```


## Factors

Factors are numeric vectors that contain only pre-defined values (categories), and where each of these categories has a label.

```{r}
a <- sample(1:2, 100, replace=T)
table(a)
```

```{r}
a_f <- factor(a, labels=c("Male","Female"))
table(a_f)
```

Note that the labels are just labels, the underlying representation is still 1 and 2.

```{r}
str(a_f)
```


Factors can sometimes cause trouble. More contemporary practice is to stick with an integer data type and add your own labels. 

```{r}
a_l <- a %>%
  set_labels(labels=c("Male","Female"))
str(a_l)
```

```{r}
frq(a)
```

```{r}
frq(a_f)
```

```{r}
frq(a_l)
```

The underlying integers behind factor labels have no ordering. To establish an ordering, make an ordered factor. 

```{r}
ord <- sample(1:5, 100, replace=T)
table(ord)
```
```{r}
ord.labs <- c("Not at all","A little","Somewhat","Much","Completely")
ord.fac <- ordered(ord, labels=ord.labs)
table(ord.fac)
```

But again, you have to be careful not to accidentally jumble the underlying integers with the ordered labels. 

To keep things more explicit, I would still stick with an integer variable with labels, rather than an ordered factor. 

```{r}
ord.l <- ord %>%
  set_labels(labels=ord.labs)
table(ord.l)

```

```{r}
frq(ord.l)
```

When you're ready to dive into this sometimes-frustrating subject, start here: 

* [forcats](https://forcats.tidyverse.org/) package in the tidyverse
* [working with labelled data](https://cran.r-project.org/web/packages/sjlabelled/vignettes/labelleddata.html)
* [wrangling categorical data in R](https://peerj.com/preprints/3163/)

## Sub-setting

You can cut up your objects into other objects. The base R way to do this is to use brackets. 

### sub-setting vectors

```{r}
a <- rpois(5, 8)
a
```

```{r}
a[1:3]
```

```{r}
a[c(1,4,2)]
```

```{r}
a[-2]
```

```{r}
a[a>9]
```

### sub-setting data frames

Since data frames are two dimensional, you usually (but not always) need to specify both dimensions.

```{r}
str(dat1)
```

```{r}

dat1_f <- dat1[1:5,2]
str(dat1_f)

```

The comma within the bracket specifies rows and columns. 

More contemporary practice uses what is referred to as the tidyverse. 

Brian's demonstration.


