---
title: "R From Zero: Session 2"
subtitle: "Breakout Groups"
author: "Brian Calhoon"
date: "5/13/2021"
output: 
  html_document:
    toc: true
    toc_depth: '5'
    toc_float: yes
---
<style>
#TOC {border-color: #133469;

}

body{font-family: Gill Sans MT;
     font-size: 18px;
     background-color: #FDFDFD;
}

h1{color: #133469; 
}

h2{color: #009CA6;
}

h3{color: #5482AB;
}

</style>
---


```{r global_options, include = FALSE}

# standard figure size and generate clean output
knitr::opts_chunk$set(fig.width=10, fig.height=8, warning=FALSE, message=FALSE, cache=TRUE, error=T)

```

```{r packages, include = FALSE}
#packages
library(extrafont)
library(extrafontdb)
library(gapminder)
library(tidyverse)
library(readxl)
```

```{r object from week 2}

dt <- read_xlsx("C:/Users/brian.calhoon/Documents/R From Zero/Week 2/datatest.xlsx")

dt
```

## Breakout Groups
There are 2 Breakout Groups 

More than 1 group can do each question in the event that we have more than 2 groups. If we have four groups, Dan and Melanie will have their groups work on Breakout Group1 and Brian and Jacob will have their groups work on Breakout Group 2. 

Breakout Group 1 = Melanie and Dan

Breakout Group 2 = Jacob and Brian

### Breakout Group 1
Question 1: Which team has the highest mean test score in Round 1? in Round 2? Which team had the largest improvement in scores? 

```{r challenge 1}
dt

dt5 <- dt %>%
  group_by(Team) %>%
  summarize(avg_score1 = mean(Score_Round1)
            , avg_score2 = mean(Score_Round2)) %>% 
  arrange(desc(avg_score1))

dt5 #Lions had the highest score in Round 1

arrange(dt5, desc(avg_score2)) #Bears had the highest score in Round 2

dt6 <- dt5 %>% 
  mutate(change_scores = avg_score2 - avg_score1) %>% 
  arrange(desc(change_scores))

dt6 #The Bears did the least bad from avg_score1 to avg_score2
```

What does the output show you?

Feel free to make plot if time.

### Breakout Group 2
Question 2: What percent of participants have "yes" for `In_person`? Use `recode()` to calculate a mean for your answer. Then, which group of `In_person`, had the highest average score for each round?

```{r yes to 1}

dt

#Recode the yeses and the nos to 0s and 1s

dt1 <- dt #this is just so we don't alter the original dataset

dt1$In_person <- recode(dt$In_person
  ,"No" = 0
  , "Yes" = 1)

dt1

#Then, take a mean of that column.
mean(dt1$In_person) #Approximately 64% of participants were In_person

#Group by the In_person variable and create summary statistics
dt_group <- dt %>%
  group_by(In_person) %>%
  summarize(avg_score1 = mean(Score_Round1)
            , avg_score2 = mean(Score_Round2))

dt_group #The Yeses scored higher the first time and the nos scored higher the second time
```


What did we do? We used `recode()` to tell R that it should change the values "No" and "Yes" to the numbers 0 and 1 in the In_person column. Note, that here we used the $ to call a specific column of the `dt` object. 

With this operation, we are not altering the entire object, only the `In_person` column. That is why we have the `dt$In_person` on the left side of the assignment operator (`<-`), and then we have it again on the right side of it.

Feel free to make a plot if time.