---
title: "R From Zero: Session 2"
subtitle: "Reading an Excel file"
author: "Brian Calhoon"
date: "4/14/2021"
output: 
  html_document:
    toc: true
    toc_depth: '5'
    toc_float: yes
---
This is the guide to walk us through week 2 of the R From 0 course. 

```{r global_options, include = FALSE}

# standard figure size and generate clean output
knitr::opts_chunk$set(fig.width=10, fig.height=8, warning=FALSE, message=FALSE, cache=TRUE, error=T)

```

```{r fonts, include = FALSE}
#These libraries are not for the class, but for myself
library(showtext)
library(gfonts)
library(ggplot2)
```

```{r formatting, include = FALSE}

# display to three decimal points, scientific notation displays at around six digits
options(digits=3, scipen=6)

# requires showtext package
showtext_auto()

# requires gfonts package
use_pkg_gfont("open-sans")

# graphic style
theme_set(theme_bw() + theme(panel.grid.minor.x=element_blank(),
                             panel.grid.minor.y=element_blank(),
                             plot.title=element_text(face="bold",size=18, hjust=.5, family = "sans"),
                             plot.subtitle = element_text(size=16, family="sans"),
                             plot.caption=element_text(size=12, family="sans"),
                             axis.title=element_text(size=16, family="sans"),
                             axis.text=element_text(size=14, family="sans"),    
                             legend.text=element_text(size=14, family="sans"),
                             strip.text=element_text(size=14, family="sans"))) 

# for single graphs, I have recently come to like borderless
nobord <- theme(panel.border=element_blank(),
                axis.ticks = element_blank())


```

## Summary of last week

Last week, we installed packages, then activated them using the library(), and then we created our first charts using the gapminder dataset.

To install the packages for the first time (and only the first time) we have to run this code, install.packages(c("gapminder", "ggplot2", "dplyr")). 

The gollowing ggplot code generates a basic chart.
  

```{r summary of last week}

library(gapminder)
library(ggplot2)
library(dplyr)

dat <- gapminder

ggplot(data = dat
       , aes(x = lifeExp, y = gdpPercap, color = continent, size = pop)) + 
  geom_point(alpha = .3)

```
Here are some simple ways to make the chart a little clearer and more appealing.

All we have to do is add two more lines of code, and the visualization starts to look a lot better.
```{r customizations}

ggplot(data = dat
       , aes(x = lifeExp, y = gdpPercap, color = continent, size = pop)) + 
  geom_point(alpha = .3)+
  scale_y_log10() +
  facet_wrap(~continent)

```

With one more line of code, we can show the linear relationship within each continent.

```{r customization 2}
ggplot(data = dat
       , aes(x = lifeExp
             , y = gdpPercap
             , color = continent
             , size = pop)) + 
  geom_point(alpha = .3)+
  scale_y_log10() +
  facet_wrap(~continent) +
  geom_smooth(method = "lm")
```
Now, onto to this week's lesson!

**Purpose:** By the end of this lesson you will have learned to read an Excel workbook in R, to understand the data types that we see, and to manipulate the data.

The libraries needed for this week are ggplot2, dplyr, and readxl. The easiest way to activate all of these is to type `library(tidyverse)`. Of course you can load them all separately as well.

Make sure that you already did the `install.packages("library")` on any libraries that you're planning to use. 

```{r libraries}

library(tidyverse)
library(readxl)
```

## Reading an Excel file
It's probably safe to say that we all work in Excel and share documents in Excel. Since it's so common, let's learn how to use R to work with an existing Excel document.

R is capable of reading data in any format that I can think of, but the user has to know what the format of the file is and where the file is stored (locally, a remote server, or online). 

We're going to use the file called "datatest.xlsx" as our example. Please save this file to your machine and copy the filepath. For example, I saved it here on my local harddrive. "C:\Users\brian.calhoon\Documents\R-0Course\Week 2\datatest.xlsx". 

It may be helpful to type `getwd()` into your Console. It will tell you the working directory of your session, and you could simply save the file there.  

In R, we have to flip the \ to a / and wrap "" around the filepath. So, it looks like this: "C:/Users/brian.calhoon/Documents/R-0Course/Week 2/datatest.xlsx". 

Type the code below, but change the filepath (everything before the "datatest.xlsx") to the correct filepath to get to where you saved the document.

```{r read dataset}

getwd()

read_xlsx("datatest.xlsx")

```

What happened? It shows the file, but there is no new object to interact with.

```{r assign .xlsx document to an object}

test <- read_xlsx("datatest.xlsx")

```

### What data types are we working with?

Let's look closer at `test`. We're going to do this in 3 ways, `glimpse()`, `str()`, and `head()`. Let's observe the differences in each of the outputs.

Same as last week.
```{r data types and structures}
glimpse(test)
```

This is base R version of examining an object's structure, and it requires no installation of a helper library.
```{r str}
str(test)
```

This is an easy way to see the first 6 rows of an object (first 6 rows is the default). 
```{r head}
head(test)

```

By adding the n argument you can adjust the number of rows that you see. This shows the first 20 rows.
```{r head 2}
head(test, n = 20)
```

We know there 5 columns with 33 rows of data. 

How is `str(test)` different? 

What are the abbreviated labels, `chr` and `dbl`? These are the data classes or data types. 
For some additional background, the primary data types that you will encounter are: 

* Logical -  `yes` or `no` or `TRUE` or `FALSE`
* Numeric -  `12.3`, `14.99`, `54`
* Integer -  `1`, `2456`, `39999`, `89`
* Double - `45.8976`, `3.14`, `.987843`
* Character - `"ball"`, `"a"`, `"2"`, `"This is a pencil."`
* Complex - `7+4x`

To look at the data class of single column we can use `class()` and use the $ symbol to call only that single row. It looks like this.
```{r class}

class(test$Name)

```

R makes it very easy to calculate summary statistics. If we want to see the min/max, mean, median, and first and third quartiles, we simply pass the object to `summary()`.

```{r summary statistics}

summary(test)

```
So, R recognized that the first 3 columns are character columns, and it cannot calculate summary statistics of these words. Instead, it tells us about the class of the column - character. It also recognized that columns 4 and 5 are numeric (dbl) and it found the min/max, mean, median, and quartiles. This gives us a fairly good sense of what is in the data.

The 3rd column, In_person, is listed as a class character, and we know from looking at the data above that it is a binary yes/no response. Another useful skill is knowing how to convert binary choices to 0s and 1s so that we can calculate summary statistics.

In R, this is quite simple.

```{r yes to 1}

test$In_person <- recode(test$In_person
  ,"No" = 0
  , "Yes" = 1)

test
```

What did we do? We used `recode()` to tell R that it should change the values "No" and "Yes" to the numbers 0 and 1 in the In_person column. Note, that here we used the $ to call a specific column of the `test` object. 

With this operation, we are not altering the entire object, only the `In_person` column. That is why we have the `test$In_person` on the left side of the assignment operator (`<-`), and then we have it again on the right side of it.

Now, let's run `summary(test)` again to see what happens.

```{r summaries again}

summary(test)

```
Now, let's explore the test scores.

## Calculate and create new columns

First, let's find the change from Round 1 to Round 2 so we can see where positive and negative change occurred. Then, let's arrange the data in descending order by the Change column.

To link these lines of code together we'll use pipes, ` %>%`.


```{r change in scores}

test2 <- test %>%
  mutate(change = Score_Round2 - Score_Round1) %>%
  arrange(desc(change))

test2
```

This created a new column called change where it shows absolute change. Oddly, only 7 scores were higher in round 2 than in round 1. Note, that I called this object test2. If I make a mistake on test2, it doesn't really affect much since test2 iteself is built on test.

Here's what happened. `mutate()` is the dplyr function to create a new column. We named the new column `change` and said that it equals the value in `Score_Round2` `-` the value in `Score_Round1`. So, it applies formula to each of the rows and shows the result in `change`.

We used `arrange(desc(test2))` to organize the data in descending order of the `change` column.

Now, let's add a column for percent change. 

```{r percent change in scores}

test3 <- test2 %>%
  mutate(percent_change = (change/Score_Round1)*100) %>%
  arrange(desc(percent_change))

test3  
```  

We called this object test3. Similar to before, we changed the object name so I know that it is built off of test2.

## Summarize data
We've created new columns, but we haven't yet reduced columns or rows. This is just as important. 

What is the average score of Score_Round1 for those who were in person? And for those were not in person?
```{r summarize}

test4 <- test3 %>%
  group_by(In_person) %>%
  summarize(Avg_score1 = mean(Score_Round1))

test4
```

In this example, we used the `group_by()` and `summarize()` syntax. You should get used to this as it's a super useful combination in R. In this example, it turns out that those who are In_person (the 1s), score nearly 4 points higher than those who were not in person. Of course, we're not testing if this is statistically significant or just noise.

With summarize, everything is deleted except the specific columns mentioned in the `group_by()` and then whatever `summarize()` is being calculated.

## Breakout Groups
There are 2 Challenges below. 

### Challenge Question 1
Challenge Question 1: Which team has the highest mean test score in Round 1? in Round 2? Which team had the highest improvement? 

```{r challenge 1}

test5 <- test3 %>%
  group_by(Team) %>%
  summarize(avg_score1 = mean(Score_Round1)
            , avg_score2 = mean(Score_Round2))

arrange(test5, desc(avg_score2))
```

What does the output show you?

### Challenge Question 2
Challenge Question 2: Which team improved the most from Round 1 to Round 2?

```{r Breakout Group 2}

test6 <- test3 %>%
  group_by(Team) %>%
  summarize(avg_change = mean(change))

arrange(test6, desc(avg_change))

```

What does the output show you?