---
title: "Dashboard Example"
format: 
  dashboard:
    orientation: columns
---

```{r}
library(ggplot2)
library(dplyr)
library(boot)
```

#  {.sidebar}

Global level side bar in ymal that will stay for all pages. Page level side page would be page level heading \## {.sidebar} or add next to row and columns. Same with orientation and scrolling. Can also add scrolling: true to ymal or in a page level if you want to have more rows and cards on one page. Can also add {.tabset} next to row to show two different cars in one column instead of two.

For code help: https://mine.quarto.pub/quarto-dashboards/2-dashboard-components

# Page 1

## Column 1

Changing the orientation to columns will make the second level headings columns instead of rows To change heights: {height="30%"} To change width: {width="80%"} \### Row 1 To change layout to filling or flowing add: {.fill} or {.flow}

### Row 2

```{r}
#| title: Highway vs. city mileage
ggplot(mpg, aes(x=cty, y=hwy)) +
  geom_point()
```

# Page 2

## Column 1

### Row - Value boxes {height="35%"}

```{r}
#| label: calculate-value

mpg <- mpg 

lowest_mileage_cty <- mpg %>% 
  filter(cty == min(cty)) %>% 
  distinct(cty) %>% 
  pull()

highest_mileage_cty <- mpg %>% 
  filter(cty == max(cty)) %>% 
  distinct(cty) %>% 
  pull()

rounded_mean_cty_mileage <- mpg %>% 
  summarize(round(mean(cty),2)) %>% 
  pull()

```

```{r}
#| content: valuebox
#| title: "Least efficient"
#| icon: fuel-pump-fill
#| color: danger
list(value = paste(lowest_mileage_cty, "mpg"))

```

```{r}
#| content: valuebox
#| title: "Most efficient"
#| icon: fuel-pump-fill
#| color: success
list(value = paste(highest_mileage_cty, "mpg"))

```


```{r}
#| content: valuebox
#| title: "Average city mileage"
#| icon: fuel-pump-fill
#| color: secondary
list(value = paste(rounded_mean_cty_mileage, "mpg"))

```

### Row 2 {height="65%"}

```{r}
n_cars <- nrow(mpg)

cat("title=", "Drive types of", n_cars, "cars")

ggplot(mpg, aes(x=drv)) +
  geom_bar()
```

```{r}
#| output: false

#add when you need to hide output

ggplot(mpg, aes(x=drv)) +
  geom_bar()
```

```{r}
#| title: One cell = one card
#| layout-ncol: 2

ggplot(mpg, aes(x=cty, y=hwy)) +
  geom_point()

ggplot(mpg, aes(x=drv)) +
  geom_bar()
```
