---
title: "Methods Corner"
subtitle: "Working with strings to cheat at Wordle"
author: "Dan Killian"
date: "1/18/2022"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: paper
    fig.caption: true
    code_folding: show
    df_print: kable
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
# standard figure size and generate clean output
knitr::opts_chunk$set(fig.height=4, fig.width=6, warning=FALSE, message=FALSE, cache=TRUE, error=T, echo=T)

library(here)

#knitr::opts_knit$set(root.dir="~/South Sudan resilience - MSI") # doesn't seem to work

source(here("../../Dropbox/r prep.R"))

```

Let's use Wordle as an excuse to practice with string manipulation. 

```{r}
dict <- data.frame(word = read.csv("Collins Scrabble Words (2019).txt", 
                       header = FALSE, skip = 2)[, 1]) %>%
                     mutate(word=tolower(word))
sample_n(dict, 100) %>%
  unlist() %>%
  unname()
```


### January 18, 2022

First word 'adieu', all grey!

⬜⬜⬜⬜⬜

Let's filter out all those vowels and the number of characters. 

```{r}
out <- dict %>%
  filter(str_detect(word, "a|e|i|u", negate=T),
         nchar(word)==5)
nrow(out)
```

1,200 words to work with 

```{r}
sample_n(out, 75) %>%
  unlist() %>%
  unname()
```


Second try 'strop'. R, o, and p yellow 

⬜⬜🟨🟨🟨

Let's filter for most common consonants, as well as the information from the previous guess. 

```{r}

out2 <- out %>%
  filter(str_detect(word, "t|n|s"),
         str_detect(word, "r|o|p"))

nrow(out2)

```

909 words to work with^[I later noticed that the sequential calls to str_detect did some additional filtering]

```{r}
sample_n(out2, 75) %>%
  unlist() %>%
  unname()
```


Third try 'poncy'. p and y green, o yellow.

🟩🟨⬜⬜🟩

In fact that's a regression since I did not choose a word with r or o. 

Let's filter for the first and last letters.

```{r}

out3 <- out2 %>%
  filter(substr(word, 1,1)=="p",
         substr(word,5,5)=="y")

nrow(out3)

```

Only 14 words possible. 

```{r}
out3 %>%
  unlist() %>%
  unname()
```

Next try pygmy, no help. 

🟩⬜⬜⬜🟩

In fact that's an even worse regression because I forgot to use a word that included o and r again. Let's filter for o and r so I won't forget, and also remove the consonants we've used so far. 

```{r}
out4 <- out3 %>%
  filter(str_detect(word, "o|r"),
         str_detect(word, "d|s|t|n|c|g|m", negate=T))
nrow(out4)
```

Only two words left. 

```{r}
out4 %>%
  unlist() %>%
  unname()
```

Tried proxy on fifth attempt, got it!

Zero credit for using any vocabulary skills (I still don't know what strop or poncy mean). But it's a good occasion to do some string manipulation. To keep some skill involved, next time I won't filter for most common consonants, but only filter out based on past guesses. 

Wordle 213 5/6

⬜⬜⬜⬜⬜
⬜⬜🟨🟨🟨
🟩🟨⬜⬜🟩
🟩⬜⬜⬜🟩
🟩🟩🟩🟩🟩


