---
title: "Figure customization in R and Stata"
subtitle: ""  
author: 
  - "Gustavo Castillo, Dan Killian"
date: "June 2021"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=5, fig.height=3, fig.retina=3,
  out.width = "80%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  fig.show = TRUE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#23395b"
)

```

```{r include=F}
library(xaringanExtra)
library(tidyverse)
library(here)
library(readxl)
library(viridis)
library(knitr)

use_panelset()

#source(here("prep.r"))

```

### Session objective

Show finished visualization in R

Narrate construction of visualization step by step (coding out loud)

Compare / contrast against figure construction in Stata

------------------------------------------------------------------------

### Visualization using a grammar of graphics

The ggplot2 package is based on a specific grammar of graphics. Under
this grammar of graphics, a visualization comprises six layers:

-   data
-   mapping (aesthetic)
-   geometry
-   facet
-   statistic
-   theme

------------------------------------------------------------------------

### Template of a data visualization

All plots require the first three layers:

-   data to visualize
-   a mapping of variables from the data to the visualization grid
    -   different types of mappings are referred to as 'aesthetics'
-   a geometry that gives a specific implementation of the specified
    aesthetic

```{r eval = F}
ggplot(data = [DATASET],
       mapping = aes(x = [X VARIABLE],
                     y = [Y VARIABLE])) +
  geom_SOMETHING() 
```

------------------------------------------------------------------------

### Example: Governance trends before and after the Arab Spring

.panelset\[\]

------------------------------------------------------------------------

class: middle

# Coding out loud

-   data
-   mapping (aesthetic)
-   geometry
-   facet
-   statistic
-   theme

------------------------------------------------------------------------

.midi\[ \> 1. **Start with the data**\] .pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. **map year to the x-axis**\]
.pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
**map Voice and Accountability Index to y-axis**\] .pull-left\[\]
.pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
map Voice and Accountability Index to y-axis \> 4. **map Country to the
color aesthetic**\] .pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. **assign a point geometry to display the data
mapping**\] .pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. **assign a line geometry to display the data mapping**\]
.pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. assign a line geometry to display the data mapping \> 7.
**assign a color blind-friendly palette**\] .pull-left\[\]
.pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. assign a line geometry to display the data mapping \> 7.
assign a color blind-friendly palette \> 8. **highlight a break in the
timeline with a vertical line**\] .pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. assign a line geometry to display the data mapping \> 7.
assign a color blind-friendly palette \> 8. highlight a break in the
timeline with a vertical line \> 9. **format the x-axis**\]
.pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3.
map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. assign a line geometry to display the data mapping \> 7.
assign a color blind-friendly palette \> 8. highlight a break in the
timeline with a vertical line \> 9. format the x-axis \> 10. **format
the y-axis**\] .pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

.midi\[ .pull-left\[ \> 1. Start with the data, \> 2. map year to the
x-axis \> 3. map Voice and Accountability Index to y-axis \> 4. map
Country to the color aesthetic \> 5. assign a point geometry to display
the data mapping \> 6. assign a line geometry to display the data
mapping \> 7. assign a color blind-friendly palette \> 8. highlight a
break in the timeline with a vertical line \> 9. format the x-axis \>
10. format the y-axis \> 11. **add end-point labels**\]\]
.pull-right\[\]

------------------------------------------------------------------------

.midi\[ .pull-left\[ \> 1. Start with the data, \> 2. map year to the
x-axis \> 3. map Voice and Accountability Index to y-axis \> 4. map
Country to the color aesthetic \> 5. assign a point geometry to display
the data mapping \> 6. assign a line geometry to display the data
mapping \> 7. assign a color blind-friendly palette \> 8. highlight a
break in the timeline with a vertical line \> 9. format the x-axis \>
10. format the y-axis \> 11. **add end-point labels**\]
.pull-right\[\]\]

| .pull-left\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3. map Voice and Accountability Index to y-axis \> 4. map Country to the color aesthetic \> 5. assign a point geometry to display the data mapping \> 6. assign a line geometry to display the data mapping \> 7. assign a color blind-friendly palette \> 8. highlight a break in the timeline with a vertical line \> 9. format the x-axis \> 10. format the y-axis \> 11. add end-point labels \> 12. **remove legend**\] .pull-right\[\]                               |
|:---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| .pull-left\[ \> 1. Start with the data, \> 2. map year to the x-axis \> 3. map Voice and Accountability Index to y-axis \> 4. map Country to the color aesthetic \> 5. assign a point geometry to display the data mapping \> 6. assign a line geometry to display the data mapping \> 7. assign a color blind-friendly palette \> 8. highlight a break in the timeline with a vertical line \> 9. format the x-axis \> 10. format the y-axis \> 11. add end-point labels \> 12. remove legend \> 13. **apply a thematic style**\] .pull-right\[\] |

.pull-left\[ \> 1. Start with the data, \> 2. map year to the x-axis \>
3. map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. assign a line geometry to display the data mapping \> 7.
assign a color blind-friendly palette \> 8. highlight a break in the
timeline with a vertical line \> 9. format the x-axis \> 10. format the
y-axis \> 11. add end-point labels \> 12. remove legend \> 13. **apply a
thematic style**\]

.pull-right\[\]

------------------------------------------------------------------------

.pull-left\[ \> 1. Start with the data, \> 2. map year to the x-axis \>
3. map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. assign a line geometry to display the data mapping \> 7.
assign a color blind-friendly palette \> 8. highlight a break in the
timeline with a vertical line \> 9. format the x-axis \> 10. format the
y-axis \> 11. add end-point labels \> 12. remove legend \> 13. **apply a
thematic style**\]

.pull-right\[\]

------------------------------------------------------------------------

.pull-left\[ \> 1. Start with the data, \> 2. map year to the x-axis \>
3. map Voice and Accountability Index to y-axis \> 4. map Country to the
color aesthetic \> 5. assign a point geometry to display the data
mapping \> 6. assign a line geometry to display the data mapping \> 7.
assign a color blind-friendly palette \> 8. highlight a break in the
timeline with a vertical line \> 9. format the x-axis \> 10. format the
y-axis \> 11. add end-point labels \> 12. remove legend \> 13. apply a
thematic style \> 14. **remove axis labels, add title and caption**\]

.pull-right\[\]

------------------------------------------------------------------------

### Stata, R concordance

How can we accomplish similar output in Stata?

```{r}

conc <- read_xlsx("Stata, R concordance.xlsx")
conc

```

### Final products

.pull-left\[\] .pull-right\[\]

------------------------------------------------------------------------

class: center, middle

# Thank you!
