---
title: "South Sudan's Conflict Context"
subtitle: "Exploring ACLED's Event Database" 
author: "Brian Calhoon"
date: "05/23/2021"
output: 
  bookdown::html_document2:
    fig.caption: T
    toc: true
    toc_depth: '5'
    toc_float: yes
    code_folding: hide
    
---
<style>
TOC {border-color: #133469;

}

body{font-family: Corbel;
     font-size: 18px;
     background-color: #FDFDFD;
}

h1{color: #133469; 
}

h2{color: #009CA6;
}

h3{color: #5482AB;
}

p.caption{
font-size: small;
}

</style>
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, cache = T, error = T)

```

```{r, include = FALSE, eval = TRUE}
#load packages
source("prep/packages_fonts_styles.R", local = knitr::knit_global())

```
# Conflict Across South Sudan

```{r download ACLED data, include = F}
#The data
events <- read_xlsx("data/SSudan.xlsx") 


events$admin1 <- events$admin1 %>% 
  recode("Western Bahr El Ghazal" = "Western Bahr el Ghazal")

#This version has a geometry column added to it
events_sf <- st_as_sf(events, coords = c("longitude", "latitude"), crs = 4326)
```

```{r, include = F}

#shape files for neighboring countries
neighbors <- geoboundaries(country = c("Uganda", "Kenya", "Sudan", "Ethiopia", "Central African Republic", "Democratic Republic of the Congo", "South Sudan"))

# Shape file for The border of South Sudan
SSudan <- geoboundaries(country = "South Sudan")

#shape files for the counties of South Sudan
states <- geoboundaries(country = "South Sudan"
                        , adm_lvl = "adm1")

#shape files for the districts of South Sudan
districts <- geoboundaries(country = "South Sudan"
                           , adm_lvl = "adm2")

SS_admin <- sf::st_join(states, SSudan)

SS_admin <- sf::st_join(districts, SS_admin)

#Download places - the ne_download #call isn't working. Using the url to #download manually.
#places <- ne_download(scale = 10
#                      , type = #"populated_places"
#                      , returnclass = #"sf")

#unzip the files and then read them 
unzip("data/ne_10m_populated_places.zip", exdir = "data/places")

places <- st_read("data/places/ne_10m_populated_places.shp")

#filter for only South Sudan
places_SS <- places %>% 
  filter(SOV0NAME == "South Sudan") %>% 
  st_as_sf(coords = c("LONGITUDE", "LATITUDE")
           , crs = 4326) 
  
places_SS$FEATURECLA <- case_when(places_SS$FEATURECLA == "Populated place" ~ "city"
                                  , places_SS$FEATURECLA != "Populated place" ~ "capital") %>%
  as.factor()
  

#download water
#water <- ne_download(scale = 10
#                     , type = #"rivers_lake_centerlines"
#                     , category = #"physical"
 #                    , returnclass = #"sf") 

#unzip rivers file and read it in
unzip("data/ne_10m_rivers_lake_centerlines.zip", exdir = "data/rivers")

water <- st_read("data/rivers/ne_10m_rivers_lake_centerlines.shp")

#filter for the region
water_SS <- water[neighbors,]

#download lakes
#lakes <- ne_download(scale = 10
#                     , type = "lakes"
#                     , category = #"physical"
#                     , returnclass = #"sf") 

#unzip lakes files and read in
unzip("data/ne_10m_lakes.zip", exdir = "data/lakes")

lakes <- st_read("data/lakes/ne_10m_lakes.shp")

```
ACLED's conflict event data from South Sudan dates from July 14, 2011 and goes through December 2, 2021. It includes 7,939 unique conflict events.

```{r, fig.width=10, fig.height=8, fig.cap = "Conflict events in South Sudan since its inception as a country."}

#A map of the country with all conflict events plotted
ggplot(data = neighbors)+
  geom_sf(fill = "#F8F0E3") +
    geom_sf_label(data = neighbors
                , aes(label = shapeName)) +
  geom_sf(data = states,
          color = "white") +
  geom_sf(data = st_jitter(events_sf, .2)
          , aes(color = event_type)
          , alpha = .3)+
  geom_sf_text(data = states
                , aes(label = shapeName)
                , color = "#565656") +
  geom_sf(data = SSudan
          , color = "#000000"
          , width = 4
          , fill = NA) +
  geom_sf(data = places_SS)+
  geom_sf_text(data = places_SS
                , aes(label = NAME_ID)
                , nudge_x = .2
                , nudge_y = -.1
                , check_overlap = T
                , color = "#000000")+
  geom_sf(data = water
          , color = "steel blue"
          , width = 4) +
  geom_sf(data = lakes
          , color = "steel blue"
          , width = 10)+
  coord_sf(xlim = c(24.15331, 35.94900)
           , ylim = c(3.248898, 12.23631))+
  theme.plot()+
  labs(title = "South Sudan"
         , subtitle = ""
         , caption = "Author: Brian Calhoon; Sources: rgeoboundaries, rnaturalearth"
       , x = ""
       , y = "")+
  theme(axis.text = element_blank())

```

```{r fig.dim = c(10, 8)}

a <-st_join(SS_admin, events_sf, left = T)

#Choropleth map of counties and conflict
a %>% 
  group_by(shapeName) %>% 
  summarize(attacks = log1p(sum(!is.na(event_id_cnty)))) %>% 
  tmap::tm_shape() +
  tmap::tm_polygons('attacks', title = 'ACLED\nevents (logged)'
                    , style = "cont"
  #, palette = cont_pal
  , border.col = "white") +
  tmap::tm_shape(states)+
  tmap::tm_borders(col = "white"
             , lwd = 2)+
  tmap::tm_shape(st_union(a)) +
  tmap::tm_borders(col = "black"
             , lwd = 2)+
  tmap::tm_layout(main.title = "Concentration of Conflict in South Sudan, 2011 - 2021"
                  , fontfamily = "Corbel"
                  , legend.outside.position = "bottom"
                  , legend.position = c(0,0)) +
  tmap::tm_credits("Source: ACLED")
```

```{r, fig.width = 10, fig.height =12, fig.cap = "Conflict varies over time across South Sudan. With events concentrating over time in Central Equatoria."}

a <- a %>%  
  group_by(shapeName, year) %>% 
  summarize(attacks = log1p(sum(!is.na(event_id_cnty))))

tmap::tmap_mode("plot")+
  tmap::tm_shape(a
                 , is.master = T) +
  tmap::tm_polygons("attacks"
                    , title = "ACLED\nevents (logged)"
                    , border.col = "grey"
                    , style = "cont") +
  tmap::tm_facets("year") + 
  tmap::tm_shape(st_union(a))+
  tmap::tm_borders(col = "black")+
  tmap::tm_layout(main.title = "Conflict Concentration by Year, 2011 - 2021"
                  , legend.outside.position = "bottom", legend.position = c(0, 0)) #.8, 1.1

```



# An interactive map
Below is a map that you can explore for yourself. If you hover over one of the dots a label will appear, and if you click on it additional information about the event will appear.

```{r, fig.cap = "An interactive map of all conflicts in South Sudan."}
#tmap::tmap_mode("view")
#tmap::tm_basemap(c(StreetMap = #"OpenStreetMap",
#             TopoMap = #"OpenTopoMap")) +
#  tmap::tm_tiles(c(TonerHybrid = #"Stamen.TonerHybrid"))+
#  tmap::tm_shape(events_sf, is.master #= TRUE) + 
#  tmap::tm_dots(col = "event_type"
#          , id = "event_type"
#          , popup.vars=c("Event date" #= "event_date"
#                         , "Sub-event #Type"="sub_event_type"
#                         , "Actors"= #"actor1"
#                         , "Actors" = #"actor2"
#                         , #"Fatalities" = "fatalities"
#                         , "Sources" #= "source")
#          , popup.format=list() 
#          , group = "Events"
#          , jitter = .2
#          , alpha = .3
#          , palette = my_pal)
```
