---
title: "BPI Baseline Preliminary Findings"
author: "Dan Killian"
date: "February 9, 2017"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo=F, message=F, warning=F)

setwd("C:/Users/dan.killian/Dropbox/work/MSI/Chevron Bangladesh (BPI)/survey/BPI Enterprise Development Baseline Report/Data")
library(haven)
library(foreign)
library(readxl)
library(sjPlot)
library(sjmisc)
library(tidyverse)
library(DescTools)
library(psych)
library(knitr)


options(digits=3, scipen=6)

theme_set(theme_bw() + theme(panel.grid.major.x=element_blank(),
                             panel.grid.minor.x=element_blank(),
                             panel.grid.major.y=element_blank(),
                             panel.grid.minor.y=element_blank(),
                             plot.title=element_text(face="bold",size=18)) + 
               theme(axis.title=element_text(size=16)) + 
               theme(axis.text=element_text(size=14)))

dat <- read.spss("Final Datafile (BPI).sav", to.data.frame=T)
dat$a20 <- dat[,1]

dat$localremitUSD <- dat$d1_res / 78
describe(dat$localremitUSD)

dat$abroadremitUSD <- dat$d2_res / 78
describe(dat$abroadremitUSD)

```
## A quick note on navigation

- 'C' Show table of contents
- 'F' Toggles the display of the footer
- 'A' Toggles display of current vs all slides (useful for printing handouts)
- 'S' Make fonts smaller
- 'B' Make fonts larger

## Bangaldesh Partnership Initiative (BPI)

Five-year, multi-sectoral initiative to improve incomes, employability, and inclusiveness in the communities surrounding Chevron gas plants. BPI objectives are to:   

- Build capacity of local enterprises to sustain and grow their business  

- Build craft vocational skills and strengthen long term employment prospects  

- Enhance the overall impact of BPI by leveraging funds and partnerships, and linking Chevron investments to regional economic development  

- Strengthen BPI effectiveness and ensure Chevron can tell a holistic and credible story of impact   


## BPI Components and Projects  

* Enterprise Development
    + Jibika (BRAC)  
  
* Workforce Development
    + Uttoron (SwissContact)
    + BACCO (British Council)  
    
* Enabling Environment
    + DIAGives Alliance (CAF America)
  
* Crosscutting
    + BPI Monitoring & Evaluation Framework (MSI)

## BPI Timeline

- 2014 Partnership launched
- Oct 2015 Enterprise Development Component launched
- Mar 2016 Crosscutting M&E Component launched  
- Aug 2016 Workforce Development Component launched
- Nov 2016 BPI Baseline Survey launched
- Jan 2017 Enabling Environment Component launched
- Oct 2017 Enterprise Development Phase 2
- Jan 2018 Workforce Development Phase 2
- Nov 2018 BPI Midline Survey
- Nov 2019 BPI Endline Survey


## Enterprise Development Component Baseline Survey  

Objectives:

- Provide an independent measure of baseline income levels, enable subsequent measurement of long term income gains

- Generate baseline measure on a broader set of 10 key socio-economic indicators

- Provide data on villages receiving other Chevron investments (schools, health clinics, etc)

- Identify characteristics of project success

## Presentation outline

- <span style="color:blue">Household respondents</span>

- Household income

- Perceptions of Chevron

## Household respondents

- The Enterprise Development Baseline survey collected household-level information from 750 respondents across Chevron's three natural gas plants. 

- Respondents are sampled evenly across the Bibiyana (Habiganj district), Jalalabad (Sylhet district), and Moulavibazar (Moulavibazar district) gas plants. 

`r sjt.xtab(dat$dist, dat$upazila, digits=1, no.output=T)$knitr`

## Household respondents

Respondents are sampled evenly across the nine BPI unions, with the exception of Sreemangal

`r sjt.frq(dat$union, digits=1,no.output=TRUE)$knitr`


## Respondent sampling 

Respondents were sampled according to two primary groups: 

- Jibika beneficiaries (300 respondents)
- Households (450 respondents)

`r sjt.xtab(dat$res_type1, dat$dist, no.output=T)$knitr`

The survey sampled from these two groups in order to distinguish between individual beneficiary and village level measures. 

## Within group sampling

Within the beneficiary group: 

    - Received seed funding (163)
    - Received microloan (48)
    - Received training, but not yet received seed funding or microloan (89)

Within the household survey: 

    - Randomly sampled households (428)
    - Randomly sampled households that happened to include a VDO member (22)

`r sjt.xtab(dat$res_type1, dat$res_type2, no.output=T)$knitr`


## Within and across village sampling

Jibika beneficiaries sampled from 50 beneficiary villages

Within each Jibika village: 

    - 6 households surveyed that participated in Jibika
    - 4 randomly selected households (which could have included a VDO member)

Within each non-Jibika village, 10 randomly selected households

Final complication: 

    - Of the 50 Jibika and 25 non-Jibika villages, 15-20 had received previous Chevron investments in schools, health clinics, or other support
    - (To be confirmed)

## Summary of design and comparison groups

```{r}

a <- c("Treatment","Villages","Households","Beneficiaries","Total respondents")
b <- c("Jibika", 50, 200, 300, 500)
c <- c("non-Jibika",25, 250, 0, 250)
d <- c("Total", 75,450,300, 750)

t <- data.frame(rbind(b,c, d), row.names=NULL)
names(t) <- a
kable(t)
```

- Note how the comparison group of 450 respondents can be divided into those in non-Jibika villages and those in Jibika villages
    - Allows test for spillover effects of Chevron programming

## Primary hypotheses and learning agenda

The Enterprise Development Baseline survey, in conjunction with follow-up surveys in 2018 and 2019 of the same respondents, will enable testing of the following hypotheses: 

    - Did Jibika support lead to long term, sustainable increase in household income?
    - Is Jibika, and BPI more broadly, contributing to poverty reduction in the communities surrounding its gas plants? 

The survey will also enable exploration of a learning agenda to gather lessons learned, unintended effects, and the characteristics of project success. 

Examples of learning agenda questions: 

    - Do improved beneficiary / VDO outcomes lead to improved village outcomes?
    - What are the individual- or village-level characteristics of success, or lack of it?
    - Are there VDO characteristics contributing to project success? 
    - Is there a separate effect from previous Chevron programming in Jibika villages? 
    - Do Jibika and previous Chevron programming interact to strengthen beneficiary and / or village outcomes? 

## Presentation outline

- Household respondents

- <span style="color:blue">Household income</span>

- Perceptions of Chevron

## Household income

Measuring household income is hard!

    - Multiple sources of income
    - Irregular / occasional labor
    - Lots of in-kind "labor" contributing to reciprocal networks of exchange based on family, neighborhood, clan, tribe
    - Informal or extra-legal labor that respondents may want hidden
    - Unclear that income is even a well-defined concept to measure among the poor

## Measuring income - approaches

Several approaches to measuring income

    - Ask directly about income sources and tolerate ambiguity and error in the measures
    - Itemize what households own - a wealth rather than income measure
    - Itemize what households spend - a consumption rather than income measure
    - Ask households to rate their economic situation - a perceptual measure of welfare

Measuring expenditures is considered the most robust measure of income, and also connects well with an economist's view of consumption as the basic measure of human welfare. 

However, measuring expenditures is an extremely detailed and exhaustive exercise that exceeded the time and resource budget of our survey. 

In this survey, we measure the following: 

    - Wage and other sources of income
    - Household assets
    - Progress out of Poverty Index (a simplified asset score linked to poverty likelihoods)
    - Self-rating of economic situation

## Primary wage income

`r sjt.frq(dat$c1_1, sort.frq="desc", no.output=T)$knitr`

## Primary other income

- 9.7 percent of households received an average of $1,080 in foreign remittances
- 3.6 percent of households received an average of $314 in domestic remittances

## Total household income (USD per day) by Subdistrict

Both wage and other income, summed across all working members of the household  

```{r}

inc <- read.csv("Total household annual income.csv")

incUpa <- read.csv("Income by Subdistrict.csv")

incUpa$dailyUSD <- round(incUpa$dailyUSD,1)
ggplot(incUpa, aes(x=dailyUSD, y=reorder(as.character(Subdistrict), dailyUSD), label=dailyUSD)) + 
  geom_point(size=3, color="darkblue") + 
  geom_segment(aes(yend=Subdistrict), xend=1, color="black") +
  geom_label(color="darkblue") + 
  xlim(1,10) + ylab("") + 
  geom_vline(xintercept=mean(incUpa$dailyUSD), color="maroon")


```


## Total household income (USD per day) by Union

Both wage and other income, summed across all working members of the household

```{r}
incUnion <- read.csv("Income by Union.csv")

ggplot(incUnion, aes(x=dailyUSD, y=reorder(as.character(Union), dailyUSD), label=dailyUSD)) + 
  geom_point(size=3, color="darkblue") + 
  geom_segment(aes(yend=Union), xend=1, color="black") +
  geom_label(color="darkblue") + 
  xlim(1,10) + ylab("") +
  geom_vline(xintercept=mean(incUnion$dailyUSD), color="maroon")



```

## Reported income and perceived economic situation 

```{r}

theme_set(theme_bw() + theme(panel.grid.major.x=element_blank(),
                             panel.grid.minor.x=element_blank(),
                             panel.grid.minor.y=element_blank(),
                             plot.title=element_text(face="bold",size=18)) + 
               theme(axis.title=element_text(size=16)) + 
               theme(axis.text=element_text(size=14)))

ladder <- inc %>%
  group_by("Perceived economic situation"=l1) %>%
  summarise(dailyUSD=mean(TotaldailyUSD))

round(cor(ladder[,1], ladder[,2]),2)

ggplot(ladder, aes(x=ladder[,1], y=ladder[,2])) + stat_smooth(se=F) + geom_point() + 
  xlab("Perceived economic situation") + ylab("Total daily income (USD)") + 
  scale_y_continuous(limits=c(0, 75), breaks=c(0,5,10,15,20,30,40,50,60,70)) + 
  scale_x_continuous(limits=c(1,9), breaks=1:9)
```

Next steps are to compute a full household asset score as well as poverty likelihoods from the Progress out of Poverty Index

## Presentation outline

- Household respondents

- Household income

- <span style="color:blue">Perceptions of Chevron</span>

## Who is aware of Chevron? 

`r sjt.xtab(dat$k1,dat$res_type1, no.output=T)$knitr`

## Perceptions of Chevron

```{r}

Desc(dat$k7, plotit=T)

```



