---
title: "Introduction to Impact Evaluation"
author: "Dan Killian"
date: "June 19, 2017"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## We want to know the causal effect of a project on its beneficiaries

Job training on earnings and employment

Class size on test scores

School leadership on teacher morale

Teacher qualifications on student outcomes

## Potential outcomes - i

Consider an indicator for a potential beneficiary, Di	

D tells us whether there is a project, or a "treatment"

The subscript i denotes a single individual who is either treated or not treated

Di = 1 ??? participates in a project (beneficiary)
Di = 0 ??? does not participate in a project


## Potential outcomes - ii

Consider an indicator for the outcome of a potential beneficiary, Yi

Y1i ??? the outcome after project participation

Y0i ??? the outcome without any project

Note that Y1 and Y0 denote *possibilities* for the *same person*, unit i ! 


## Potential outcomes - iii

The effect of the project (treatment effect) on person i is the difference between the two potential outcomes
Treatment effect = Y1i - Y0i 

This is the difference in potential outcomes for the *same person*

A person participates in a project, and then goes back in time and does not participate in the project


## Potential outcomes - iv

But how can one person be both treated and untreated?

In the real world, person i experiences one of the potential outcomes, but not both

If Di = 1, the potential outcome of Yi becomes Y1i
If Di = 0, the potential outcome of Yi becomes Y0i


## The fundamental problem of causal inference

This is the fundamental problem of causal inference:

We observe only one outcome, but we need both outcomes to describe the effect of the project

We refer to the outcome that didn't happen as the *counterfactual*, or what would have happened in the absence of the project


```{r}

Group <- c("Treatment","Control")
Y1i <- c(<font color="red">"Observed"</font color>, "Counterfactual")


```



























